---
title: 'Data 607: Project 2, Part I'
author: "Aaron Grzasko"
date: "October 9, 2016"
output: 
    html_document:
        theme: simplex
        highlight: haddock
---

---

## Assignment Overview

The purpose of this exercise is to gain additional experience manipulating untidy data, using data sets suggested by students in CUNY's Data 607 class.  
  
In Part I, I use a data set submitted by my fellow classmate, Jose Zuniga.  
  
The source of the original data set can be found on the website [databasejournal.com](http://www.databasejournal.com/features/mysql/article.php/3861886/Tips-for-Simplifying-Crosstab-Query-Statements.htm)

Jose's original discussion post on this data can be found [here](https://bbhosted.cuny.edu/webapps/discussionboard/do/message?action=list_messages&forum_id=_1397022_1&nav=discussion_board&conf_id=_1377703_1&course_id=_1310339_1&message_id=_20871555_1#msg__20871555_1Id).  
  
  
## Load Libraries  
  
```{r, warning = FALSE, message=FALSE}
# uncomment script below to install necessary packages
# install.packages(c("tidyr","dplyr","stringr","car"))

library(tidyr)
library(dplyr)
library(stringr)
library(car)
library(knitr)
```
  
## Data
  
The data set for this exercise contains monthly citizen counts for 9 separate months and 5 regions within a presumably fictional country.  
  
The csv file containing this data is in a wide format, with the variable "Region" spanning multiple columns.  
  
Let's load the data:  
  
```{r, comment=NA}
myurl <- "https://raw.githubusercontent.com/spitakiss/Data607_Project2_1/master/citizen.csv"
citizen <- read.csv(myurl, header = TRUE, stringsAsFactors = FALSE)
citizen <- tbl_df(citizen)
citizen
```
  
## Tidy Operations  
  
1. *Create month name-to-number function*.  It is usually more practical and efficient to store dates as numbers.  I created a custom function, `month_num`, to handle the conversion of the raw month name to number format.  Note:  I used a recommended approach from stack overflow to create this function--see [stackoverflow](http://stackoverflow.com/questions/6987478/convert-a-month-abbreviation-to-a-numeric-month-in-r).  
  
2. *Convert table from wide format to long format*.  As mentioned previously, we wish to collapse the *region* variable into one column, as this format is often easier to work with for analytical purposes.  

  
3. *Remove TOTAL row*.  The TOTAL row is superfluous; so I remove it using the `filter` function.  If at a later time we need to calculate these totals, we can do so using a host of different functions in base R and other packages.    

4. *Clean values for region variable*.  In this step, I remove the string prefix, "REGION" from all *region* values.  This step is accomplished using the `mutate()` function and `str_replace_all()` function from the stringr package.  
  
5. *Implement `month_num()` function*.  See step 1 for more details.  
  
6. *Ignore TOTAL Column*  This is step is similar to step 3.  The TOTAL column is superfluous and is removed using the `select()` function.  


```{r, comment=NA}
# 1
month_num <- function(x){
   x <- tolower(x)
   month_list <- list("january" = 1, "february" = 2, "march" = 3,"april"= 4, "may" = 5, "june" = 6, "july" = 7, "august" = 8 ,
                      "september" = 9, "october" = 10, "november" = 11, "december" = 12)
   sapply(x,function(x) month_list[[x]])
}


citizen <- citizen %>%
                # 2
                gather("region","num_citizen", REGION1:REGION1.4) %>%
                # 3
                filter(MONTH != 'TOTAL') %>%
                mutate(
                # 4
                    region = str_replace_all(region,pattern = "REGION", replacement = ""),
                # 5   
                     MONTH = as.numeric(month_num(MONTH))
                )%>% 
                # 6    
                select(-TOTAL) %>%
                rename(month = MONTH)
head(citizen)
```
  
## Analysis  
  
The instructions for analysis provided by Jose were fairly open-ended:  
  
> "Compare monthly citizenship for the given regions.""


**Graphs**  
Let's start with a graph that plots monthly citizen counts by month and region.  
  
Below I use a function from the *car package*, `scatterplot()`.  This function works well with long format data, and requires a minimal amount of coding effort.  
  

```{r graph}
# plot each region in one graph
scatterplot(citizen$num_citizen ~ citizen$month | citizen$region, smoother = FALSE, reg.line = FALSE,  xlab = "Month", ylab = "Citizens", main = "Plot Ct by Region")    

# sum citizens in every region
total_cit <- citizen %>%
                select(month,num_citizen) %>%
                group_by(month) %>%
                summarise(sum = sum(num_citizen))

# plot total citzenship by month
scatterplot(total_cit$sum ~ total_cit$month, smoother = FALSE, reg.line = FALSE, col = "blue",  xlab = "Month", ylab = "Citizens", main = "Plot Ttl Ct")
```
  
It is clear from these plots that total citizenship spiked considerably from April through August and stayed relatively flat throughout the year thereafter.  
  
We also see that Region 1.2 experienced the greatest spike in population, while Regions 1 and 1.3 remained sparsely populated for the entire 9 month period.  
  
  
**Summary Statistics**  
  
In the tables below I calculate basic summary statistics:  
  
* Average citizenship by region for the entire 9 month period  
* Ending citizenship by region for December, the last month in the data set  

```{r}
average_stat <- citizen %>%
                select(region, num_citizen) %>%
                group_by(region) %>%
                summarise(avg_citizens = round(mean(num_citizen),1)) %>%
                arrange(desc(avg_citizens))
knitr::kable(average_stat)


ending_stat <- citizen %>%
    rename(dec_citizens = num_citizen) %>%
    filter(month == 12) %>%
    select(region, dec_citizens) %>%
    arrange(desc(dec_citizens))


knitr::kable(ending_stat)

```
  
From these summary statistics, we see that all regions ended the year with populations above their respective 9 month averages, with the exception of regions 1.3 and 1, which experienced a decrease.  

